---
title: "6. Generalized linear mixed models"
format: html
execute: {echo: true, warning: false, message: false}
---

```{r}
#| label: setup
#| include: false
source("_common.R")
library(dplyr)
library(ggplot2)
library(lme4)
```

## 1. Aim

GLMMs estimate subject-specific effects for non-Gaussian repeated outcomes by conditioning on latent random effects. Likelihood inference integrates over the random-effects distribution; Laplace and adaptive Gauss--Hermite quadrature trade computation against approximation accuracy.

## 2. Model specification

$$
g\{E(Y_{ij}\mid b_i)\}=x_{ij}^{\mathsf T}\beta+z_{ij}^{\mathsf T}b_i+log(e_{ij}),
\qquad b_i\sim N(0,D),
$$
with conditional independence of $Y_{ij}$ given $b_i$. The marginal likelihood contribution is
$$
L_i(\beta,D)=\int\left\{\prod_j f(y_{ij}\mid b_i;\beta)\right\}\phi(b_i;0,D)\,db_i.
$$

## 3. R packages/functions

`lme4::glmer()` fits standard-family GLMMs using Laplace (`nAGQ = 1`) or adaptive Gauss--Hermite quadrature (`nAGQ > 1`, scalar random effect). `glmmTMB::glmmTMB()` supports negative-binomial, zero-inflated, hurdle, beta-binomial, offset, and dispersion submodels. `DHARMa` supplies simulation-based diagnostics.

## 4. Data

`MASS::epil` contains four repeated seizure counts per subject; `exposure = 2` denotes the two-week count interval.

```{r}
data("epil", package = "MASS")
epil <- MASS::epil |>
  mutate(subject = factor(subject), period_c = period - 1, exposure = 2)
```

## 5. Fit

```{r}
form <- y ~ trt * period_c + lbase + lage + offset(log(exposure)) + (1 | subject)

pois_laplace <- glmer(
  form, data = epil, family = poisson(link = "log"), nAGQ = 1,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)
pois_agq <- update(pois_laplace, nAGQ = 7)

summary(pois_laplace)
cbind(Laplace = fixef(pois_laplace), AGQ7 = fixef(pois_agq))
AIC(pois_laplace, pois_agq)
```

```{r}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  nb_glmm <- glmmTMB::glmmTMB(
    y ~ trt * period_c + lbase + lage + offset(log(exposure)) + (1 | subject),
    data = epil,
    family = glmmTMB::nbinom2(link = "log")
  )
  summary(nb_glmm)
  AIC(pois_laplace, nb_glmm)
}
```

### Transition model

```{r}
epil_transition <- epil |>
  group_by(subject) |>
  arrange(period, .by_group = TRUE) |>
  mutate(lag_y = lag(y)) |>
  ungroup() |>
  filter(!is.na(lag_y))

transition_fit <- glm(
  y ~ trt + period_c + log1p(lag_y) + lbase + lage,
  family = quasipoisson(link = "log"), data = epil_transition
)
summary(transition_fit)
```

## 6. Extract & interpret

```{r}
exp(cbind(estimate = fixef(pois_agq), confint(pois_agq, parm = "beta_", method = "Wald")))
as.data.frame(VarCorr(pois_agq))
head(ranef(pois_agq)$subject)

conditional_rate <- predict(pois_agq, type = "response", re.form = NULL)
fixed_re_zero_rate <- predict(pois_agq, type = "response", re.form = NA)
head(cbind(conditional_rate, fixed_re_zero_rate))
```

Fixed-effect exponentials are subject-specific conditional rate ratios at equal random effects. The random-intercept variance is heterogeneity on the log-rate scale; `ranef()` returns empirical-Bayes conditional modes. `re.form = NULL` predicts for fitted subjects conditional on their random effects; `re.form = NA` sets random effects to zero and is not, for a nonlinear link, the marginal population mean.

Under a logistic random-intercept model,
$$
\beta_{PA}\approx\beta_{SS}/\sqrt{1+0.35\sigma_b^2},
$$
so the corresponding GEE coefficient is typically attenuated. Under a log link with normal random intercept, marginalization adds $\sigma_b^2/2$ to the intercept when random-effect variance is covariate-invariant; slopes then retain the same multiplicative interpretation.

## 7. Diagnostics

```{r}
lme4::isSingular(pois_laplace)
pois_laplace@optinfo$conv$lme4$messages

pearson <- residuals(pois_laplace, type = "pearson")
overdispersion <- sum(pearson^2) / (nobs(pois_laplace) - length(fixef(pois_laplace)))
overdispersion

ggplot(data.frame(fitted = fitted(pois_laplace), pearson), aes(fitted, pearson)) +
  geom_point() + geom_hline(yintercept = 0, linetype = 2)

if (requireNamespace("DHARMa", quietly = TRUE)) {
  sim_res <- DHARMa::simulateResiduals(pois_laplace, n = 250, plot = FALSE)
  plot(sim_res)
  DHARMa::testDispersion(sim_res)
  DHARMa::testZeroInflation(sim_res)
}
```

Compare Laplace and higher-order quadrature when cluster sizes are small or random-effect variance is large; inspect gradients/Hessian, singularity, simulated residual uniformity, dispersion, zero inflation, and subject-level influence.

## 8. Caveats

- GEE is population-averaged; GLMM is subject-specific. Do not compare coefficient magnitudes without marginalization and link-specific attenuation.
- `glmer()` uses Laplace at `nAGQ = 1`; `nAGQ = 0` is a faster, less accurate approximation. Adaptive quadrature with `nAGQ > 1` is restricted to a single scalar random-effect term.
- SAS `PROC GLIMMIX` may default to pseudo-likelihood or different quadrature depending on options; match the estimation method explicitly before comparing results.
- Overdispersion can reflect omitted random effects, serial dependence, zero inflation, or a wrong mean/variance family; a negative-binomial fit is not an automatic cure.
- LRTs for random-effect variances have boundary nulls; plain chi-square p-values can be anticonservative.
- Conditional modes are shrunken predictions and their empirical distribution is not a direct normality test of the latent population.
- A transition-model coefficient conditions on observed response history and is neither a GEE marginal effect nor a random-effects subject-specific effect; initial-condition and missing-history handling must be stated.
- **Source-backed:** conditional interpretation, marginal likelihood, quadrature/approximations, and empirical Bayes follow the notes. **Standard-practice addition:** `glmmTMB` and `DHARMa` supply current idiomatic R implementations.

## 9. Source references

- `/Users/wenbinwu/Documents/UNC/26Spring/BIOS767/Lecture-Notes/BIOS-767-Part-015.pdf`, pp. 5--24, 31--39, and 50--61 (GLMM specification, estimation, interpretation).
- `/Users/wenbinwu/Documents/UW/25Winter/STAT571/notes/6.0 - GLMM_1.pdf`, pp. 1--16 and 22--30 (subject-specific models).
- `/Users/wenbinwu/Documents/UW/25Winter/STAT571/notes/6.1 - GLMM_2.pdf`, pp. 5--13 (GEE/GLMM marginalization and attenuation).
- `/Users/wenbinwu/Documents/UW/25Winter/STAT571/notes/6.2 - GLMM_3.pdf`, pp. 1--21; `/Users/wenbinwu/Documents/UW/25Winter/STAT571/notes/6.3 - GLMM_4.pdf`, pp. 1--7 (quadrature, Laplace, PQL, empirical Bayes).
- `/Users/wenbinwu/Documents/UNC/26Spring/BIOS767/Assignments/UNC-BIOS767/analysis_Wenbin/Assignment7/BIOS767-HW7.Rmd`, code chunks `fit_models` and `random_effects_predictions` (R GLMM comparison and predictions).
- `/Users/wenbinwu/Documents/UW/25Winter/STAT571/HW/HW5/STAT571-HW5-Code.Rmd`, GLMM/GEE comparison chunks (conditional versus marginal fits).
- `/Users/wenbinwu/Documents/UNC/26Spring/BIOS767/Lecture-Notes/BIOS-767-Part-020.pdf`, pp. 2--32 (transition models and lagged responses).
