---
title: "7. Missing longitudinal data"
format: html
execute: {echo: true, warning: false, message: false}
---

::: {.callout-note title="Mathematical companion"}
[Missing longitudinal data: mechanisms, likelihood, EM, MI, IPW, and MNAR sensitivity](math-writeups/10-missing-data.qmd)
:::

```{r}
#| label: setup
#| include: false
source("_common.R")
library(dplyr)
library(tidyr)
library(lme4)
```

## 1. Aim

Missing-data analysis targets the full-data estimand under an explicit observation-process assumption. Likelihood/MI methods are generally valid under MAR given an adequate joint or imputation model; ordinary GEE typically requires MCAR, whereas inverse-probability-weighted GEE can be valid under correctly modeled MAR dropout. MNAR requires an unverifiable selection, pattern-mixture, shared-parameter, or sensitivity model.

## 2. Model specification

Let $R_{ij}=1$ indicate observation. Mechanisms are
$$
\begin{aligned}
\text{MCAR: }&P(R\mid Y_{obs},Y_{mis},X)=P(R\mid X),\\
\text{MAR: }&P(R\mid Y_{obs},Y_{mis},X)=P(R\mid Y_{obs},X),\\
\text{MNAR: }&P(R\mid Y_{obs},Y_{mis},X)\text{ depends on }Y_{mis}.
\end{aligned}
$$
MI combines $m$ estimates with
$$
\bar Q=m^{-1}\sum Q_k,\quad T=\bar U+(1+m^{-1})B.
$$
For monotone dropout, IPW solves $\sum_i D_i^{\mathsf T}V_i^{-1}W_i(Y_i-\mu_i)=0$, where $W_i$ contains inverse estimated observation probabilities.

## 3. R packages/functions

`mice` performs multiple imputation and Rubin pooling; wide-format imputation is often convenient for a small fixed visit grid. `geepack` accepts observation/dropout weights. `tidyr::fill()` demonstrates LOCF but does not make it valid. `naniar` visualizes patterns.

## 4. Data

An incomplete version of `nlme::Orthodont` is generated reproducibly with monotone dropout after age 10 depending on observed age-10 distance and sex; this is MAR by construction.

```{r}
data("Orthodont", package = "nlme")
orth <- as.data.frame(Orthodont) |>
  arrange(Subject, age) |>
  mutate(Subject = factor(Subject), male = as.integer(Sex == "Male"), age_c = age - 8)

set.seed(767)
drop_info <- orth |>
  filter(age == 10) |>
  transmute(
    Subject,
    p_drop = plogis(-0.8 + 0.35 * (distance - mean(distance)) + 0.35 * male),
    dropout = rbinom(n(), 1, p_drop)
  )

orth_miss <- orth |>
  left_join(drop_info, by = "Subject") |>
  mutate(distance_obs = ifelse(dropout == 1 & age >= 12, NA_real_, distance))
with(orth_miss, table(age, is.na(distance_obs)))
```

## 5. Fit

### LOCF (display only; not the primary analysis)

```{r}
locf_data <- orth_miss |>
  group_by(Subject) |>
  mutate(distance_locf = distance_obs) |>
  fill(distance_locf, .direction = "down") |>
  ungroup()
head(locf_data[is.na(locf_data$distance_obs), c("Subject", "age", "distance_obs", "distance_locf")])
```

### Multiple imputation in wide visit format

```{r}
wide_miss <- orth_miss |>
  select(Subject, male, age, distance_obs) |>
  pivot_wider(names_from = age, values_from = distance_obs, names_prefix = "d") |>
  mutate(Subject = as.integer(Subject))

if (requireNamespace("mice", quietly = TRUE)) {
  ini <- mice::mice(wide_miss, maxit = 0, printFlag = FALSE)
  meth <- ini$method
  pred <- ini$predictorMatrix
  meth[c("Subject", "male")] <- ""
  pred[, "Subject"] <- 0
  pred["male", ] <- 0

  imp <- mice::mice(
    wide_miss, m = 10, maxit = 10,
    method = meth, predictorMatrix = pred,
    seed = 767, printFlag = FALSE
  )

  fits <- lapply(seq_len(imp$m), function(k) {
    mice::complete(imp, k) |>
      pivot_longer(starts_with("d"), names_to = "age", values_to = "distance") |>
      mutate(age = as.numeric(sub("d", "", age)), age_c = age - 8, Subject = factor(Subject)) |>
      lmer(distance ~ male * age_c + (1 | Subject), data = _, REML = FALSE)
  })
  mi_pool <- mice::pool(mice::as.mira(fits))
  summary(mi_pool, conf.int = TRUE)
}
```

### Observed-data EM estimate under multivariate normality

```{r}
if (requireNamespace("norm", quietly = TRUE)) {
  y_wide <- as.matrix(wide_miss[c("d8", "d10", "d12", "d14")])
  prelim <- norm::prelim.norm(y_wide)
  theta_em <- norm::em.norm(prelim, showits = FALSE)
  em_parameters <- norm::getparam.norm(prelim, theta_em)
  em_parameters$mu
  em_parameters$sigma
}
```

### Pattern-mixture sensitivity parameterization

```{r}
pattern_fit <- lmer(
  distance_obs ~ male + age_c * factor(dropout) + (1 | Subject),
  data = orth_miss, REML = FALSE
)
summary(pattern_fit)
```

### Dropout model and weighted GEE

```{r}
ipw_dat <- orth_miss |>
  group_by(Subject) |>
  mutate(
    visit = row_number(),
    R = as.integer(!is.na(distance_obs)),
    lag_y = lag(distance_obs),
    at_risk = lag(R, default = 1L) == 1L
  ) |>
  ungroup()

den_fit <- glm(
  R ~ male + age_c + lag_y,
  family = binomial,
  data = filter(ipw_dat, visit > 1, at_risk)
)
num_fit <- glm(
  R ~ male + age_c,
  family = binomial,
  data = filter(ipw_dat, visit > 1, at_risk)
)

risk_rows <- which(ipw_dat$visit > 1 & ipw_dat$at_risk)
ipw_dat$p_den <- 1
ipw_dat$p_num <- 1
ipw_dat$p_den[risk_rows] <- pmax(predict(den_fit, type = "response"), 0.02)
ipw_dat$p_num[risk_rows] <- pmax(predict(num_fit, type = "response"), 0.02)
ipw_dat <- ipw_dat |>
  group_by(Subject) |>
  mutate(sw = cumprod(ifelse(R == 1, p_num / p_den, 1))) |>
  ungroup()

if (requireNamespace("geepack", quietly = TRUE)) {
  wg <- geepack::geeglm(
    distance_obs ~ male * age_c,
    id = Subject, waves = visit, weights = sw,
    data = filter(ipw_dat, R == 1),
    family = gaussian, corstr = "independence"
  )
  summary(wg)
}
```

## 6. Extract & interpret

MI estimates combine across completed datasets; `ubar`, `b`, and total variance distinguish within- and between-imputation uncertainty. IPW coefficients target the marginal mean in the pseudo-population induced by inverse observation probabilities. The dropout logistic coefficients describe association with continued observation, not the scientific outcome model.

## 7. Diagnostics

```{r}
if (exists("imp")) {
  plot(imp)                         # chain means/SDs by iteration
  mice::densityplot(imp, ~d14)
}

summary(ipw_dat$sw[ipw_dat$R == 1])
quantile(ipw_dat$sw[ipw_dat$R == 1], c(0, .01, .5, .99, 1))
table(ipw_dat$R, ipw_dat$visit)
```

For MI, inspect convergence, observed/imputed distributions, transformations, bounds, interactions, and whether the imputation model is at least as rich as the analysis model. For IPW, inspect positivity, extreme weights, dropout-model calibration, and sensitivity to truncation.

## 8. Caveats

- MCAR/MAR/MNAR are statements about the joint full-data/observation process; observed data alone cannot distinguish MAR from MNAR.
- Likelihood ignorability under MAR also requires distinct parameters and correct outcome modeling. MI requires congenial imputation/analysis models and inclusion of predictors of missingness and outcome.
- LOCF treats an imputed value as known, distorts trajectories and covariance, and is generally unjustified except under a scientifically defensible constant-after-dropout estimand.
- Standard `mice` wide imputation is appropriate for a modest common visit grid; irregular or multilevel data may require two-level methods (`2l.pan`, `2l.norm`) or joint modeling.
- Ordinary GEE is not generally valid under MAR dropout. IPW requires a correct observation model and positivity; weighted GEE scale and covariance defaults differ across R/SAS implementations.
- Selection and pattern-mixture models are not identified under MNAR without restrictions. Report sensitivity parameters (e.g., delta adjustment/tipping point), not a single unverifiable fit.
- EM maximizes the observed-data likelihood under its joint model; it does not by itself add between-imputation uncertainty or make an MNAR mechanism ignorable.
- **Source-backed:** taxonomy, likelihood ignorability, MI, IPW, selection/pattern-mixture models, and LOCF limitations follow the notes. **Standard-practice addition:** `mice`/`geepack` code is the native R implementation.

## 9. Source references

- `BIOS-767-Part-018-1.pdf`, pp. 8--13, 21--36, 44--62, and 69--87 (mechanisms, likelihood, MI, weighting).
- `BIOS-767-Part-019.pdf`, pp. 4--18 and 24--45 (missing-data examples, MI and weighting).
- `missing_data_writeup.pdf`, pp. 1--8 (likelihood ignorability under MCAR/MAR/MNAR).
- UW STAT571 (missingness mechanisms, selection and pattern-mixture models, and weighted GEE).
- BIOS767 Homework 4 (patterns and dropout prediction).
- `BIOS-767-Part-004.pdf`, pp. 4--33 (missing longitudinal data and EM algorithm).
