---
title: "A1. Inference and diagnostics"
format: html
execute: {echo: true, warning: false, message: false}
---

::: {.callout-note title="Mathematical companion"}
[Inference and diagnostics](math-writeups/05-inference-diagnostics.qmd)
:::

```{r}
#| label: setup
#| include: false
source("_common.R")
library(dplyr)
library(ggplot2)
library(lme4)
library(nlme)
```

## 1. Aim

This appendix consolidates inferential and diagnostic operations by model class. Diagnostics target mean/link specification, residual covariance, random-effects adequacy, cluster influence, numerical convergence, and distributional dispersion; no single residual definition diagnoses all components.

## 2. Model specification

For an LMM, marginal residuals are $r_i^{(m)}=Y_i-X_i\widehat\beta$ and conditional residuals are $r_i^{(c)}=Y_i-X_i\widehat\beta-Z_i\widehat b_i$. For GLM/GLMM mean $\widehat\mu$, Pearson residuals are $(Y-\widehat\mu)/\sqrt{\widehat{\operatorname{Var}}(Y)}$; deviance residuals are signed square roots of unit-deviance contributions. Cluster influence compares $\widehat\beta$ with $\widehat\beta_{(-i)}$.

## 3. R packages/functions

`residuals()` exposes model-specific residuals; `nlme::Variogram()` assesses remaining lag dependence; `lme4::isSingular()` and optimizer output assess fitted random-effects geometry; `influence.ME` supplies cluster deletion; `DHARMa` supplies simulation-based GLMM diagnostics; `performance` offers convenient standardized checks.

## 4. Data

`Orthodont` supplies a Gaussian LMM example and `MASS::epil` supplies a repeated-count GLMM example.

```{r}
data("Orthodont", package = "nlme")
orth <- as.data.frame(Orthodont) |>
  arrange(Subject, age) |>
  mutate(Subject = factor(Subject), age_c = age - 8, visit = ave(age, Subject, FUN = seq_along))

data("epil", package = "MASS")
epil <- MASS::epil |>
  mutate(subject = factor(subject), period_c = period - 1)
```

## 5. Fit

```{r}
lmm <- lmer(distance ~ Sex * age_c + (age_c | Subject), orth, REML = TRUE)
gls_ar1 <- gls(
  distance ~ Sex * age_c, orth,
  correlation = corAR1(form = ~visit | Subject), method = "REML"
)
count_glmm <- glmer(
  y ~ trt * period_c + lbase + lage + (1 | subject),
  epil, family = poisson, nAGQ = 1
)
```

## 6. Extract & interpret

```{r}
if (requireNamespace("lmerTest", quietly = TRUE)) {
  lmm_lt <- lmerTest::as_lmerModLmerTest(lmm)
  anova(lmm_lt, ddf = "Satterthwaite")
  anova(lmm_lt, ddf = "Kenward-Roger")
}

if (requireNamespace("pbkrtest", quietly = TRUE)) {
  lmm_reduced <- lmer(distance ~ Sex + age_c + (age_c | Subject), orth, REML = TRUE)
  pbkrtest::KRmodcomp(lmm, lmm_reduced)
}

confint(lmm, parm = "theta_", method = "profile")
```

Fixed-effect tests require the intended estimable function and denominator-df method. Variance-component CIs should respect the zero boundary. For nested fixed-effect LMM comparisons refit under ML unless using a KR comparison specifically formulated for REML fits with common random structure.

## 7. Diagnostics

### Gaussian marginal and conditional residuals

```{r}
lmm_diag <- data.frame(
  fitted_conditional = fitted(lmm),
  residual_conditional = resid(lmm),
  residual_pearson = resid(lmm, type = "pearson"),
  fitted_marginal = predict(lmm, re.form = NA),
  residual_marginal = orth$distance - predict(lmm, re.form = NA)
)

ggplot(lmm_diag, aes(fitted_conditional, residual_pearson)) +
  geom_point() + geom_hline(yintercept = 0, linetype = 2)
qqnorm(lmm_diag$residual_conditional); qqline(lmm_diag$residual_conditional)

re <- ranef(lmm)$Subject
qqnorm(re[["(Intercept)"]]); qqline(re[["(Intercept)"]])
qqnorm(re[["age_c"]]); qqline(re[["age_c"]])
```

### Serial dependence and influence

```{r}
vg <- Variogram(gls_ar1, form = ~age | Subject, resType = "normalized")
plot(vg, smooth = TRUE)

if (requireNamespace("influence.ME", quietly = TRUE)) {
  infl <- influence.ME::influence(lmm, group = "Subject")
  cd <- cooks.distance(infl)
  head(sort(cd, decreasing = TRUE))
  dfbetas(infl)[1:5, , drop = FALSE]
}
```

### GLMM residuals, dispersion, and convergence

```{r}
glmm_diag <- data.frame(
  fitted = fitted(count_glmm),
  pearson = residuals(count_glmm, type = "pearson"),
  deviance = residuals(count_glmm, type = "deviance")
)
ggplot(glmm_diag, aes(fitted, pearson)) + geom_point() + geom_hline(yintercept = 0, lty = 2)

sum(glmm_diag$pearson^2) / (nobs(count_glmm) - length(fixef(count_glmm)))
isSingular(count_glmm)
count_glmm@optinfo$conv$lme4$messages

if (requireNamespace("DHARMa", quietly = TRUE)) {
  dh <- DHARMa::simulateResiduals(count_glmm, n = 250, plot = FALSE)
  plot(dh)
  DHARMa::testDispersion(dh)
  DHARMa::testZeroInflation(dh)
}

if (requireNamespace("performance", quietly = TRUE) &&
    requireNamespace("see", quietly = TRUE)) {
  performance::check_model(count_glmm)
}
```

## 8. Caveats

- Marginal residuals retain random-effect variation; conditional residuals depend on estimated BLUPs and can mask random-effect misspecification.
- QQ plots of BLUPs do not directly test the latent random-effects distribution because BLUPs are heteroscedastic, correlated, and shrunken.
- Pearson/deviance residuals are discrete and nonnormal in GLMMs; simulation-based residuals are usually more interpretable.
- Cook's distance and DFBETAs should be deleted by independent cluster, not by row, for longitudinal data.
- A flat residual variogram is evidence against remaining second-order lag structure only after the mean and random effects are adequate.
- Singularity is a boundary estimate, not an optimizer failure; convergence warnings require gradient/Hessian and rescaling checks before changing optimizers.
- SAS `PROC MIXED` influence measures and R package measures can use different scaling; compare rankings and deleted estimates, not labels alone.
- **Source-backed:** residual taxonomy, variograms, and influence follow BIOS 767/STAT 571. **Standard-practice addition:** `DHARMa`, `performance`, and `influence.ME` provide native R diagnostics.

## 9. Source references

- `BIOS-767-Part-012.pdf`, pp. 5--14 and 26--50 (marginal/conditional residuals, variograms, influence).
- UW STAT571 (serial correlation, residuals, and random-effect checks).
- `BIOS_767_Weideman_Typed_Notes_with_Examples.pdf`, pp. 86--92 (cumulative residuals and cluster diagnostics).
- BIOS767 Homework 4 (observation/case deletion influence and KR inference).
- BIOS767 Homework 5 (residual, QQ, and influence code).
