---
title: "A3. R packages for longitudinal data analysis"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note title="Mathematical companion"}
[Math-writeup catalogue: map each model family to its formulation, assumptions, and inference](math-writeups/00-index.qmd)
:::

## 1. Aim

This appendix maps common longitudinal-data tasks to idiomatic R packages. Package choice is driven by the estimand, response distribution, dependence model, and inferential target; no single package covers all combinations.

## 2. Model specification

The package groups cover conditional mixed models
$$
g\{E(Y_{ij}\mid b_i)\}=x_{ij}^{\mathsf T}\beta+z_{ij}^{\mathsf T}b_i,
\qquad b_i\sim N(0,D),
$$
and marginal models
$$
g\{E(Y_{ij})\}=x_{ij}^{\mathsf T}\beta,
\qquad \operatorname{Var}(Y_i)=A_i^{1/2}R_i(\alpha)A_i^{1/2}.
$$

## 3. R packages/functions

### Data manipulation and visualization

- **dplyr** — long-format grouping, summaries, joins, and creation of subject/time variables.
- **tidyr** — wide-to-long reshaping with `pivot_longer()` and balanced-grid construction with `complete()`.
- **ggplot2** — individual profiles, mean-profile plots, residual displays, and diagnostic graphics.
- **naniar** — missingness summaries and pattern visualizations such as `vis_miss()` and `gg_miss_upset()`.

### Gaussian repeated measures and linear mixed models

- **nlme** — `gls()`/`lme()` models with explicit residual covariance structures, including CS, AR(1), continuous-time AR(1), spatial correlation, and unstructured correlation.
- **lme4** — sparse, modular fitting of linear and generalized linear mixed models via `lmer()` and `glmer()`.
- **lmerTest** — Satterthwaite or Kenward--Roger denominator-df tests for `lmer()` fixed effects.
- **pbkrtest** — Kenward--Roger and parametric-bootstrap comparisons for nested Gaussian mixed models.

### Marginal models and GEE

- **geepack** — `geeglm()` GEE fitting, robust/model-based covariance extraction, common working correlations, and QIC via `QIC()`.
- **gee** — alternative GEE implementation whose summary exposes naive and robust standard errors and scale estimation directly.

### Generalized linear mixed models

- **lme4** — Laplace or adaptive Gauss--Hermite fitting of standard-family GLMMs through `glmer(..., nAGQ=)`.
- **glmmTMB** — flexible GLMMs with offsets, zero inflation, dispersion models, and families such as negative binomial and beta-binomial.

### Missing data and weighting

- **mice** — multiple imputation by chained equations, passive transformations, multilevel imputation methods, and Rubin pooling.
- **mitools** — manipulation and pooling of multiply imputed analyses when model-specific `mice::pool()` support is unavailable.
- **WeightIt** — construction and diagnostics of inverse-probability weights; longitudinal dropout weights usually require explicit visit-specific models.
- **survey** — weighted estimating-equation analyses with design-aware sandwich variance estimation.

### Contrasts, predictions, and inference

- **emmeans** — estimated marginal means, estimable functions, interaction contrasts, trends, and multiplicity adjustments.
- **marginaleffects** — response-scale predictions, contrasts, slopes, and marginal standardization across model classes.
- **multcomp** — simultaneous general linear hypotheses and multiplicity-adjusted confidence intervals/tests.

### Diagnostics and influence

- **performance** — standardized checks for singularity, convergence, collinearity, residual behavior, and overdispersion.
- **see** — plotting backend used by `performance::check_model()` diagnostic panels.
- **DHARMa** — simulation-based residual diagnostics for GLMMs, including dispersion and zero-inflation tests.
- **influence.ME** — cluster-deletion influence measures and Cook's distances for `lme4` mixed models.
- **clubSandwich** — cluster-robust covariance estimators and small-sample corrections for supported fitted-model classes.

### Tidying and reproducible output

- **broom** — tidy coefficient, fitted-value, and residual output for GLMs and many marginal models.
- **broom.mixed** — tidy fixed effects, random effects, variance components, and confidence intervals from mixed models.
- **modelsummary** — compact, reproducible comparison tables across supported model classes.

## 4. Data

`nlme::Orthodont` contains repeated orthodontic distance measurements in long format: one row per subject-age occasion, with subject (`Subject`), time (`age`), sex, and continuous response (`distance`).

```{r}
data("Orthodont", package = "nlme")
str(nlme::Orthodont)
```

## 5. Fit

The following audit is intentionally non-failing: it identifies which manual dependencies are available before rendering the analysis chapters.

```{r}
manual_packages <- c(
  "dplyr", "tidyr", "ggplot2", "naniar",
  "nlme", "lme4", "lmerTest", "pbkrtest",
  "geepack", "gee", "glmmTMB", "mice", "mitools", "WeightIt", "survey",
  "emmeans", "marginaleffects", "multcomp",
  "performance", "see", "DHARMa", "influence.ME", "clubSandwich",
  "broom", "broom.mixed", "modelsummary"
)

package_audit <- data.frame(
  package = manual_packages,
  installed = vapply(manual_packages, requireNamespace, logical(1), quietly = TRUE)
)
package_audit

# Install once if needed:
# install.packages(package_audit$package[!package_audit$installed])
```

## 6. Extract & interpret

```{r}
transform(
  subset(package_audit, installed),
  version = vapply(package, function(p) as.character(utils::packageVersion(p)), character(1))
)
```

Package versions are part of the computational specification because supported covariance structures, inferential defaults, and tidiers can change between releases.

## 7. Diagnostics

```{r}
stopifnot(
  getRversion() >= "4.2.0",
  requireNamespace("nlme", quietly = TRUE)
)
```

Before an exam or reproducible render, run the package audit and record `sessionInfo()`; diagnose discrepancies against the fitted model's package version rather than assuming cross-package defaults agree.

## 8. Caveats

- `lme4::lmer()` deliberately omits denominator-df p-values; use `lmerTest` or `pbkrtest` when finite-sample fixed-effect tests are required.
- `nlme` residual covariance structures and `lme4` random-effects structures are not interchangeable parameterizations of every marginal covariance model.
- GEE packages differ in scale estimation, finite-sample corrections, and the labels used for robust versus model-based covariance estimates.
- SAS `PROC MIXED`/`PROC GLIMMIX` defaults are procedure- and option-dependent; reproduce the intended likelihood, denominator-df method, covariance parameterization, and quadrature explicitly in R.
- **Standard-practice addition:** packages not used in the course code are included only when they supply a current, idiomatic R implementation of a method covered in the notes.

## 9. Source references

- `BIOS_767_Weideman_Typed_Notes_with_Examples.pdf`, pp. 42--90 (mixed models, GEE/GLMM, interpretation, diagnostics).
- `BIOS-767-Part-010.pdf`, pp. 5--73 (`PROC MIXED` functionality mapped to `nlme`/`lme4`).
- `BIOS-767-Part-014.pdf`, pp. 39--80 (GEE); `BIOS-767-Part-015.pdf`, pp. 5--61 (GLMM).
- UW STAT571 (LMM, GEE, GLMM, and missing-data package examples).
- BIOS767 homework materials and UW STAT571 (R package use in course assignments; SAS inference and covariance options are reproduced with native R in the topic chapters).
