---
title: "A0. Setup: packages and exam data"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

This page is the one-time setup reference for a clean R installation. It installs the packages referenced by the manual and the available applied-exam solutions, then shows the exact data-loading patterns used in those solutions. Run the installation chunk interactively; it is not evaluated while the site renders.

## 1. Install all required packages

The manual package list includes packages used in executable chapters and packages catalogued in [A3. R packages for longitudinal data analysis](A3-packages.qmd). The exam list covers the available 2019, 2020, 2021, 2023, 2024, and 2025 applied solutions.

```{r}
#| eval: false
# Manual dependencies: data work, models, inference, missingness, and diagnostics.
manual_packages <- c(
  "broom", "broom.mixed", "car", "clubSandwich", "DHARMa",
  "dplyr", "emmeans", "gee", "geepack", "ggplot2", "glmmTMB",
  "influence.ME", "lme4", "lmerTest", "marginaleffects", "MASS",
  "mice", "mitools", "modelsummary", "multcomp", "naniar", "nlme",
  "norm", "ordinal", "pbkrtest", "performance", "see", "survey",
  "tidyr", "WeightIt"
)

# Exam-solution dependencies include document rendering, Excel import, and prediction methods.
exam_solution_packages <- c(
  "car", "e1071", "geepack", "glmnet", "glmmTMB", "knitr",
  "lme4", "MASS", "mice", "nlme", "randomForest", "readxl", "rmarkdown"
)

# unique() removes overlap; sort() makes the installation list easy to inspect.
required_packages <- sort(unique(c(manual_packages, exam_solution_packages)))

# installed.packages() reports the current libraries; setdiff() keeps only missing packages.
installed_packages <- rownames(installed.packages())
packages_to_install <- setdiff(required_packages, installed_packages)

if (length(packages_to_install) > 0L) {
  # Install runtime dependencies from CRAN; omit Suggests to avoid unrelated optional packages.
  install.packages(
    packages_to_install,
    repos = "https://cloud.r-project.org",
    dependencies = c("Depends", "Imports", "LinkingTo")
  )
} else {
  message("All required packages are already installed.")
}

# requireNamespace() verifies installation without attaching every package and creating name conflicts.
package_available <- vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)

if (!all(package_available)) {
  stop(
    "Still unavailable: ",
    paste(required_packages[!package_available], collapse = ", ")
  )
}
```

Restart R after installation. Then load only the packages needed by the chapter or solution you are running:

```{r}
#| eval: false
# library(package) attaches its exported functions for the current R session.
library(dplyr)
library(ggplot2)
library(lme4)
```

Some packages contain compiled code. On macOS or Linux, install the current compiler toolchain if R must build a package from source. On Windows, use the matching Rtools version when a binary package is unavailable.

## 2. Locate an exam's data directory

Each solution first sources `_manual_bridge.R`. The bridge finds the qualifying-exam repository root, sources the manual helpers, and makes `repo_root` available. This avoids a machine-specific absolute path.

```{r}
#| eval: false
# current_input(dir = TRUE) returns the rendered solution file's path; dirname() gets its folder.
solution_dir <- dirname(
  normalizePath(knitr::current_input(dir = TRUE), mustWork = TRUE)
)

# source(file) runs the shared bridge, which defines repo_root and loads manual helpers.
source(file.path(solution_dir, "_manual_bridge.R"))

# file.path() joins path components using the current operating system's separator.
data_path <- file.path(
  repo_root,
  "resources",
  "previous-exams",
  "DoctoralApplication-2009-present",
  "2025aug" # replace with 2019aug, 2020aug, 2021aug, 2023aug, or 2024aug
)

# Stop immediately if the year directory is wrong, then display its available files.
stopifnot(dir.exists(data_path))
list.files(data_path)
```

When working interactively outside an R Markdown solution, define `repo_root` once from the local checkout and keep all later paths relative to it:

```{r}
#| eval: false
# Replace the placeholder with the qualifying-exam repository directory on this computer.
repo_root <- normalizePath("<PATH_TO_QUALIFY_EXAM>", mustWork = TRUE)
```

## 3. Load comma-separated files with `read.csv()`

`read.csv()` is the main reader in every available solution. It assumes a header row, comma separators, and periods as decimal marks.

```{r}
#| eval: false
# read.csv(file, check.names): file is the CSV path; FALSE preserves original column names.
hiv <- read.csv(
  file.path(data_path, "HIV_data_clean.csv"),
  check.names = FALSE
)
```

Use `check.names = FALSE` when the exact source column names matter. With the default `check.names = TRUE`, R converts non-syntactic names to valid R names. Add `na.strings = c("", "NA")` only when the file's missing-value coding requires it.

## 4. Load whitespace-delimited files with `read.table()`

The 2019 and 2020 solutions use `read.table()` for `.dat` and `.txt` files whose fields are separated by spaces or tabs.

```{r}
#| eval: false
# read.table(file, header): header = TRUE means the first record contains variable names.
illness <- read.table(
  file.path(data_path, "illness1.dat"),
  header = TRUE
)
```

By default, `read.table()` treats one or more whitespace characters as separators. Set `sep` explicitly if inspection shows a different delimiter.

## 5. Load semicolon-separated files with `read.csv2()`

The 2024 student file is semicolon-separated and begins with a UTF-8 byte-order mark.

```{r}
#| eval: false
# read.csv2() defaults to sep = ";"; dec sets the decimal mark; fileEncoding handles the BOM.
student_raw <- read.csv2(
  file.path(data_path, "student.csv"),
  dec = ".",
  check.names = FALSE,
  fileEncoding = "UTF-8-BOM"
)
```

Ordinary `read.csv2()` defaults to `dec = ","`; the solution uses `dec = "."` because this particular file stores decimal points.

## 6. Load Excel workbooks with `readxl::read_excel()`

The 2025 solution reads the HIV codebook from an `.xlsx` workbook. `read_excel()` returns a tibble, so `as.data.frame()` is useful when later code expects a base data frame.

```{r}
#| eval: false
library(readxl)

# read_excel(path, sheet): path identifies the workbook; sheet selects a name or one-based position.
hiv_codebook <- as.data.frame(
  readxl::read_excel(
    file.path(data_path, "HIV_data_codebook.xlsx"),
    sheet = 1
  )
)
```

## 7. Reader map for the available solutions

| Exam solution | Data-loading functions |
|---|---|
| 2019 | `read.csv()`, `read.table()` |
| 2020 | `read.csv()`, `read.table()` |
| 2021 | `read.csv()` |
| 2023 | `read.csv()` |
| 2024 | `read.csv()`, `read.csv2()` |
| 2025 | `read.csv()`, `readxl::read_excel()` |

No available solution uses `load()` or `readRDS()`. Do not substitute those functions unless the supplied data are actually `.RData` or `.rds` files.

## 8. Check data immediately after loading

```{r}
#| eval: false
# These checks confirm dimensions, names, types, representative rows, and missing-value counts.
dim(hiv)
names(hiv)
str(hiv)
head(hiv)
colSums(is.na(hiv))
```

Always inspect the exam PDF and its same-year codebook before recoding variables. A successful file import does not establish that factor levels, missing-value codes, units, or observational structure were interpreted correctly.
