---
title: "A2. Interpretation cheatsheet"
format: html
execute: {echo: true, warning: false, message: false}
---

::: {.callout-note title="Mathematical companions"}
[Math-writeup catalogue](../math-writeups/00-index.qmd) · [Gaussian marginal models](../math-writeups/03-gaussian-marginal-models.qmd) · [LMMs](../math-writeups/04-linear-mixed-models.qmd) · [GEE](../math-writeups/07-marginal-models-gee.qmd) · [GLMMs](../math-writeups/08-generalized-linear-mixed-multilevel.qmd) · [Transition models](../math-writeups/11-transition-models.qmd)
:::

```{r}
#| label: setup
#| include: false
source("_common.R")
library(geepack)
library(emmeans)
library(marginaleffects)
library(MASS)
library(dplyr)
library(lme4)
```

## 1. Aim

This appendix maps common output quantities to their estimands and valid one-sentence interpretations. Interpretations are conditional on coding, link, time origin, interaction structure, covariance estimator, and missing-data assumptions.

## 2. Model specification

$$
\text{GEE: }g\{E(Y_{ij})\}=x_{ij}^{\mathsf T}\beta_{PA},
\qquad
\text{GLMM: }g\{E(Y_{ij}\mid b_i)\}=x_{ij}^{\mathsf T}\beta_{SS}+z_{ij}^{\mathsf T}b_i.
$$
For Gaussian LMMs, identity-link marginalization leaves $\beta$ unchanged. For logistic random-intercept models, $\beta_{PA}\approx\beta_{SS}/\sqrt{1+0.35\sigma_b^2}$.

## 3. R packages/functions

Use `summary()` for the complete fitted object; `broom.mixed::tidy()` for fixed/random parameters; `emmeans` for estimable means/trends/interaction contrasts; `marginaleffects` for response-scale standardization; `VarCorr()`/`ranef()` for variance components and BLUPs.

## 4. Data

`MASS::epil` is converted to a binary repeated outcome for a direct population-average versus subject-specific comparison.

```{r}
data("epil", package = "MASS")
epil <- MASS::epil |>
  # subject is the cluster ID; period_c centers time; any_seizure converts count y to binary.
  mutate(subject = factor(subject), period_c = period - 1, any_seizure = as.integer(y > 0)) |>
  arrange(subject, period)
```

## 5. Fit

```{r}
# geeglm(formula, id, waves, data, family, corstr): fit a population-average binary model.
gee_bin <- geepack::geeglm(
  any_seizure ~ trt * period_c + lbase + lage,
  # subject clusters rows; period orders waves; exchangeable uses one within-subject correlation.
  id = subject, waves = period, data = epil,
  family = binomial, corstr = "exchangeable"
)

# glmer(): same response/fixed mean, plus (1 | subject) for subject-specific random intercepts.
glmm_bin <- lme4::glmer(
  any_seizure ~ trt * period_c + lbase + lage + (1 | subject),
  # binomial uses the logit link by default; nAGQ = 7 controls likelihood integration accuracy.
  epil, family = binomial, nAGQ = 7
)
```

## 6. Extract & interpret

```{r}
# vbeta is geeglm's robust sandwich covariance; diagonal square roots are robust SEs.
cbind(estimate = coef(gee_bin), robust_se = sqrt(diag(gee_bin$geese$vbeta)))
# Exponentiating logit fixed effects gives conditional odds ratios.
exp(fixef(glmm_bin))
as.data.frame(VarCorr(glmm_bin))

# emtrends(model, specs, var): estimate period_c trends by treatment on the requested response scale.
ss_trends <- emmeans::emtrends(glmm_bin, ~trt, var = "period_c", type = "response")
ss_trends
# pairs() compares treatment-specific trends; Holm adjustment controls multiplicity.
pairs(ss_trends, adjust = "holm")

# avg_comparisons(model, variables, type) averages the response-probability treatment contrast over rows.
marginaleffects::avg_comparisons(glmm_bin, variables = "trt", type = "response")
```

| Quantity | Concise interpretation |
|---|---|
| LMM fixed intercept | Expected response for the reference covariate pattern at the time origin; recenter time if this is extrapolative. |
| LMM fixed slope | Expected response change per time unit for the reference group; with an interaction it is not the common slope. |
| Group x time | Difference in time slopes (continuous time) or difference of occasion contrasts (factor time). |
| Random-intercept variance | Residual between-subject heterogeneity in the conditional intercept. |
| Random-slope variance | Residual between-subject heterogeneity in conditional change; marginal variance/correlation then depends on time. |
| Random-effect correlation | Association between latent deviations; inspect time centering because it changes the intercept and its correlation with slope. |
| BLUP/conditional mode | Shrunken subject deviation conditional on the fitted population model and variance parameters. |
| Random-intercept ICC | $\tau_0^2/(\tau_0^2+\sigma^2)$ for the Gaussian random-intercept model only. |
| GEE coefficient | Population-average change in the link of the marginal mean. |
| GEE working $\alpha$ | Nuisance association parameter used for efficiency; not generally a target unless association is explicitly modeled. |
| GEE robust SE | Sandwich uncertainty, asymptotically robust to working-correlation misspecification. |
| GEE model-based SE | Valid only if the working covariance is correctly specified. |
| GEE scale $\phi$ | Marginal variance multiplier under the package's scale-estimation rule. |
| Logistic GLMM $e^{\beta}$ | Subject-specific conditional odds ratio at equal random effects. |
| Logistic GEE $e^{\beta}$ | Population-average marginal odds ratio. |
| Poisson/log GLMM $e^{\beta}$ | Conditional rate ratio; an offset fixes the exposure denominator. |
| Satterthwaite/KR df | Approximate denominator df for a specified fixed-effect test; KR also adjusts covariance/test statistic. |

## 7. Diagnostics

Before interpreting, verify reference levels, time units/origin, contrast matrix, link/response scale, robust versus model-based covariance, convergence/singularity, dispersion, missing-data assumption, and whether predictions are conditional (`re.form = NULL`), random-effects-zero (`re.form = NA`), or marginalized/standardized.

```{r}
# contrasts() reveals treatment coding; isSingular()/optinfo diagnose the random-effect fit.
contrasts(epil$trt)
isSingular(glmm_bin)
glmm_bin@optinfo$conv$lme4$messages
```

## 8. Caveats

- A main effect involved in an interaction is evaluated at the reference/zero value of the interacting variable.
- `re.form = NA` in a nonlinear mixed model sets $b=0$; it does not integrate over $b$. Use marginal standardization or numerical integration for population-average predictions.
- Odds ratios are noncollapsible; PA/SS differences can occur without confounding and are not explained solely by bias.
- `emmeans` averages over the reference grid according to its weighting rule; state whether weights are equal, proportional, or standardized to a target population.
- Report effect, scale, unit, contrast, uncertainty method, and multiplicity adjustment in the same sentence.
- **Source-backed:** interpretation distinctions and attenuation follow BIOS 767/STAT 571. **Standard-practice addition:** `marginaleffects` gives explicit response-scale standardization.

## 9. Source references

- UW STAT571 (conditional and marginal model families, PA/SS relationships, and interpretation).
- `BIOS_767_Weideman_Typed_Notes_with_Examples.pdf`, pp. 80--85 (generalized mixed-model coefficient interpretation and attenuation).
- `BIOS-767-Part-010.pdf`, pp. 40--58 and 62--71 (fixed effects, random effects, BLUPs, ICC).
- `BIOS-767-Part-014.pdf`, pp. 39--52 and 62--68 (marginal GEE interpretation and covariance estimators).
