---
title: "Data manipulation and tables"
format: html
execute: {echo: true, warning: false, message: false}
---

::: {.callout-tip title="Exam pattern"}
Applied exams repeatedly ask for descriptive tables overall and by group, counts and proportions, missing-data summaries, and combined results across outcomes or studies. The verbs below cover the common preparation steps.
:::

```{r}
#| label: setup
#| include: false
source("_common.R")
library(dplyr)
library(tidyr)
```

## 1. Example data

The examples use built-in data so this page remains runnable. One row represents one participant.

```{r}
exam_dat <- mtcars |>
  tibble::rownames_to_column("car") |>
  as_tibble() |>
  transmute(
    id = row_number(),
    car,
    treatment = factor(am, labels = c("Control", "Intervention")),
    site = factor(cyl),
    mpg,
    weight = wt,
    biomarker = if_else(id %% 7 == 0, NA_real_, qsec),
    event = factor(if_else(vs == 1, "Yes", "No"), levels = c("No", "Yes"))
  )

# glimpse() checks names, types, and representative values.
glimpse(exam_dat)
```

## 2. Select, filter, arrange, and de-duplicate

```{r}
analysis_dat <- exam_dat |>
  # select() keeps variables; rename() changes names without changing values.
  select(id, treatment, site, outcome = mpg, weight, biomarker, event) |>
  # relocate() changes display order only.
  relocate(event, .after = treatment) |>
  # filter() keeps rows satisfying the stated condition.
  filter(weight < 5) |>
  # arrange() sorts rows; desc() requests decreasing order.
  arrange(treatment, desc(outcome))

analysis_dat |>
  # distinct() keeps one row per key; use .keep_all = TRUE to retain other columns.
  distinct(id, .keep_all = TRUE) |>
  slice_head(n = 6)
```

Useful row filters:

```{r}
exam_dat |> filter(is.na(biomarker))                    # missing biomarker
exam_dat |> filter(!is.na(biomarker), event == "Yes")  # complete and event-positive

exam_dat |>
  # if_any() is TRUE when at least one selected variable is missing.
  filter(if_any(c(mpg, weight, biomarker), is.na))

exam_dat |>
  # if_all() is TRUE only when every selected variable satisfies the condition.
  filter(if_all(c(mpg, weight, biomarker), ~ !is.na(.x)))
```

## 3. Create and recode variables

```{r}
exam_dat <- exam_dat |>
  mutate(
    # case_when() is readable for mutually exclusive categories.
    weight_group = case_when(
      weight < 2.5 ~ "Low",
      weight < 3.5 ~ "Middle",
      TRUE ~ "High"
    ),
    weight_group = factor(weight_group, levels = c("Low", "Middle", "High")),
    # if_else() is type-safe for a two-way definition.
    complete_biomarker = if_else(is.na(biomarker), "Missing", "Observed"),
    # across() applies the same transformation to several variables.
    across(c(mpg, weight), ~ as.numeric(scale(.x)), .names = "z_{.col}")
  )

exam_dat |> select(id, weight_group, complete_biomarker, starts_with("z_")) |> slice_head(n = 6)
```

For model reference levels, use `relevel(factor(x), ref = "Control")`. Use `lag()` or `lead()` only after sorting within subject:

```{r}
orth_lag <- nlme::Orthodont |>
  as_tibble() |>
  group_by(Subject) |>
  arrange(age, .by_group = TRUE) |>
  mutate(previous_distance = lag(distance)) |>
  ungroup()
```

## 4. Grouped summaries

### Continuous variables

```{r}
continuous_summary <- exam_dat |>
  group_by(treatment) |>
  summarize(
    n = n(),
    mean_mpg = mean(mpg, na.rm = TRUE),
    sd_mpg = sd(mpg, na.rm = TRUE),
    median_biomarker = median(biomarker, na.rm = TRUE),
    q1_biomarker = quantile(biomarker, 0.25, na.rm = TRUE),
    q3_biomarker = quantile(biomarker, 0.75, na.rm = TRUE),
    missing_biomarker = sum(is.na(biomarker)),
    .groups = "drop"  # return an ungrouped result
  )
continuous_summary
```

Use `across()` when the same statistics are needed for several variables:

```{r}
exam_dat |>
  group_by(treatment) |>
  summarize(
    across(
      c(mpg, weight, biomarker),
      list(mean = ~ mean(.x, na.rm = TRUE), sd = ~ sd(.x, na.rm = TRUE)),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )
```

### Counts and proportions

```{r}
proportion_table <- exam_dat |>
  # count() is shorthand for group_by(...) |> summarize(n = n()).
  count(treatment, event, name = "n") |>
  group_by(treatment) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()
proportion_table
```

For denominators based on distinct participants, use `n_distinct(id)` rather than `n()`.

### Missingness

```{r}
missing_summary <- exam_dat |>
  summarize(
    across(
      everything(),
      list(n_missing = ~ sum(is.na(.x)), pct_missing = ~ mean(is.na(.x))),
      .names = "{.col}_{.fn}"
    )
  ) |>
  pivot_longer(
    everything(),
    names_to = c("variable", ".value"),
    names_pattern = "(.*)_(n_missing|pct_missing)"
  )
missing_summary
```

## 5. Reshape with `tidyr`

Long format is usually best for models and `ggplot2`; wide format is useful for paired variables, covariance matrices, and some multivariate procedures.

```{r}
orth_long <- nlme::Orthodont |>
  as_tibble() |>
  mutate(Subject = factor(Subject), age = as.integer(age))

orth_wide <- orth_long |>
  # pivot_wider() creates one response column per age.
  pivot_wider(
    id_cols = c(Subject, Sex),
    names_from = age,
    values_from = distance,
    names_prefix = "age_"
  )

orth_long_again <- orth_wide |>
  # pivot_longer() gathers the age columns into key-value pairs.
  pivot_longer(
    starts_with("age_"),
    names_to = "age",
    names_prefix = "age_",
    values_to = "distance",
    values_drop_na = TRUE
  ) |>
  mutate(age = as.integer(age))
```

Expose absent subject-visit rows before diagnosing missingness:

```{r}
orth_balanced <- orth_long |>
  slice(-c(2, 8)) |>
  group_by(Subject) |>
  # complete() adds absent scheduled ages as explicit rows.
  complete(age = sort(unique(orth_long$age))) |>
  # fill() restores the subject-level variable on newly created rows.
  fill(Sex, .direction = "downup") |>
  ungroup()
```

`drop_na()` performs complete-case filtering. Use it only when that analysis is justified; do not silently replace a missing outcome with zero.

## 6. Join and stack data

```{r}
site_key <- tibble(
  site = factor(c(4, 6, 8)),
  site_label = c("Clinic A", "Clinic B", "Clinic C")
)

exam_labeled <- exam_dat |>
  # left_join() keeps every analysis row and appends matching lookup fields.
  left_join(site_key, by = "site")

trial_1 <- exam_dat |> mutate(trial = "Trial 1")
trial_2 <- exam_dat |> mutate(trial = "Trial 2", id = id + max(exam_dat$id))

# bind_rows() stacks datasets with compatible columns.
combined_trials <- bind_rows(trial_1, trial_2)
```

After any join, compare row counts and check key uniqueness. An accidental many-to-many join can duplicate participants.

## 7. Basic `knitr::kable()` output

```{r}
continuous_summary |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  knitr::kable(
    caption = "Descriptive statistics by treatment group",
    col.names = c(
      "Treatment", "N", "Mean MPG", "SD MPG", "Median biomarker",
      "Q1 biomarker", "Q3 biomarker", "Missing biomarker"
    ),
    align = c("l", rep("r", 7))
  )
```

For percentages, create a display column before printing:

```{r}
proportion_table |>
  mutate(percent = sprintf("%.1f%%", 100 * proportion)) |>
  select(Treatment = treatment, Event = event, N = n, Percent = percent) |>
  knitr::kable(caption = "Event frequency by treatment", align = c("l", "l", "r", "r"))
```

`kable()` formats an already prepared table; do calculations and rounding upstream. A concise exam table should have an informative caption, readable labels, units where relevant, and the denominator behind every percentage.

## 8. High-yield function map

| Task | Functions |
|---|---|
| Inspect | `glimpse()`, `names()`, `dim()`, `slice_head()` |
| Keep or order | `select()`, `filter()`, `arrange()`, `distinct()` |
| Create or recode | `mutate()`, `if_else()`, `case_when()`, `across()` |
| Summarize | `group_by()`, `summarize()`, `count()`, `n()`, `n_distinct()` |
| Reshape | `pivot_longer()`, `pivot_wider()`, `complete()`, `fill()` |
| Combine | `left_join()`, `inner_join()`, `bind_rows()` |
| Print | `knitr::kable()` |

## 9. Exam references

- 2019 Applied Doctoral Exam, pp. 5 and 8--9: preliminary summaries, response-by-subtype proportions, and combined displays.
- 2020 Applied Doctoral Exam, pp. 4--7: descriptive tables, sample sizes, response rates, confidence intervals, and multi-study summaries.
- 2021 Applied Doctoral Exam, pp. 2, 4, and 9--10: overall and group-specific descriptive tables.
- 2023--2025 Applied Doctoral Exams: descriptive tables by group, missingness summaries, predictor summaries, and publication-ready model tables recur throughout.
