---
title: "Common ggplot2 visualizations"
format: html
execute: {echo: true, warning: false, message: false}
---

::: {.callout-tip title="Exam pattern"}
Applied exams commonly request clear EDA, outcome prevalence by group, longitudinal trajectories, model-based comparisons, and graphical diagnostics. Start with one question per figure and label the estimand and units.
:::

```{r}
#| label: setup
#| include: false
source("_common.R")
library(dplyr)
library(tidyr)
library(ggplot2)
```

## 1. Example data and grammar

```{r}
exam_dat <- mtcars |>
  tibble::rownames_to_column("car") |>
  as_tibble() |>
  transmute(
    id = row_number(),
    treatment = factor(am, labels = c("Control", "Intervention")),
    site = factor(cyl),
    mpg,
    weight = wt,
    event = factor(if_else(vs == 1, "Yes", "No"), levels = c("No", "Yes"))
  )
```

Every plot has the same structure:

```r
ggplot(data, aes(x = x_variable, y = y_variable)) +
  geom_*() +
  labs(x = "Readable x label", y = "Readable y label")
```

Put variables inside `aes()` and fixed choices outside it: `aes(color = treatment)` maps a variable, while `color = "navy"` sets one color.

## 2. One-variable distributions

### Continuous outcome

```{r}
# binwidth controls resolution; choose it from the outcome scale.
ggplot(exam_dat, aes(x = mpg)) +
  geom_histogram(binwidth = 3, boundary = 0, color = "white", fill = "steelblue") +
  labs(x = "Fuel efficiency (miles per gallon)", y = "Number of participants")
```

```{r}
# A boxplot summarizes center/spread; jitter shows the actual observations.
ggplot(exam_dat, aes(x = treatment, y = mpg, fill = treatment)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.08, height = 0, alpha = 0.7) +
  guides(fill = "none") +
  labs(x = NULL, y = "Fuel efficiency (miles per gallon)")
```

### Categorical outcome

```{r}
# geom_bar() computes counts from raw rows.
ggplot(exam_dat, aes(x = event)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Event", y = "Number of participants")
```

## 3. Counts and proportions by group

```{r}
# position = "fill" converts each treatment bar to within-group proportions.
ggplot(exam_dat, aes(x = treatment, fill = event)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(x = NULL, y = "Participants", fill = "Event")
```

Use precomputed values with `geom_col()`:

```{r}
event_summary <- exam_dat |>
  count(treatment, event, name = "n") |>
  group_by(treatment) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()

# geom_col() uses the supplied y values; dodge places groups side by side.
ggplot(event_summary, aes(treatment, proportion, fill = event)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(x = NULL, y = "Event prevalence", fill = "Event")
```

Always state whether percentages are within treatment, site, visit, or the full sample.

## 4. Association between continuous variables

```{r}
# geom_smooth(method = "lm") overlays group-specific linear fits and confidence bands.
ggplot(exam_dat, aes(weight, mpg, color = treatment)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Weight (1,000 lb)", y = "Fuel efficiency (miles per gallon)", color = "Treatment")
```

Use a log scale for a positive, strongly right-skewed variable when scientifically meaningful:

```r
plot_object + scale_x_log10()
```

## 5. Longitudinal profiles

```{r}
orth <- nlme::Orthodont |>
  as_tibble() |>
  mutate(Subject = factor(Subject))

# group identifies trajectories; facet_wrap() creates comparable panels.
ggplot(orth, aes(age, distance, group = Subject, color = Sex)) +
  geom_line(alpha = 0.4, linewidth = 0.5) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ Sex) +
  guides(color = "none") +
  labs(x = "Age (years)", y = "Orthodontic distance")
```

Summarize the trajectories when individual lines are too dense:

```{r}
mean_profile <- orth |>
  group_by(Sex, age) |>
  summarize(
    n = sum(!is.na(distance)),
    mean = mean(distance, na.rm = TRUE),
    se = sd(distance, na.rm = TRUE) / sqrt(n),
    .groups = "drop"
  )

# Error bars are pointwise mean +/- SE, not confidence intervals unless multiplied by a critical value.
ggplot(mean_profile, aes(age, mean, color = Sex, group = Sex)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  labs(x = "Age (years)", y = "Mean distance (mean +/- SE)", color = "Sex")
```

## 6. Estimates and confidence intervals

```{r}
effect_table <- tibble(
  predictor = factor(c("Treatment", "Female", "Site B", "Site C")),
  estimate = c(0.78, 1.24, 0.92, 1.41),
  conf_low = c(0.61, 0.95, 0.70, 1.03),
  conf_high = c(0.99, 1.62, 1.21, 1.93)
)

# coord_flip() gives long predictor labels more room.
ggplot(effect_table, aes(predictor, estimate)) +
  geom_hline(yintercept = 1, linetype = 2, color = "grey45") +
  geom_pointrange(aes(ymin = conf_low, ymax = conf_high)) +
  coord_flip() +
  labs(x = NULL, y = "Odds ratio (95% CI)")
```

For differences, the null line is 0; for odds, risk, and rate ratios, it is 1.

## 7. Model-based predictions

```{r}
fit <- lm(mpg ~ weight * treatment, data = exam_dat)

weight_support <- exam_dat |>
  group_by(treatment) |>
  summarize(min_weight = min(weight), max_weight = max(weight), .groups = "drop")
common_weight_range <- c(max(weight_support$min_weight), min(weight_support$max_weight))

prediction_grid <- expand_grid(
  # Restrict comparisons to the weight range observed in both groups.
  weight = seq(common_weight_range[1], common_weight_range[2], length.out = 60),
  treatment = levels(exam_dat$treatment)
)

prediction_grid <- bind_cols(
  prediction_grid,
  as_tibble(predict(fit, newdata = prediction_grid, interval = "confidence"))
)

# The ribbon shows uncertainty in the fitted mean, not prediction intervals for individuals.
ggplot(prediction_grid, aes(weight, fit, color = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.18, color = NA) +
  geom_line(linewidth = 0.9) +
  labs(
    x = "Weight (1,000 lb)", y = "Predicted mean fuel efficiency",
    color = "Treatment", fill = "Treatment"
  )
```

Plot adjusted predictions from the fitted model when the question asks about an adjusted effect; raw group means answer a different question.

## 8. Diagnostics

```{r}
diagnostic_dat <- tibble(
  fitted = fitted(fit),
  residual = residuals(fit)
)

# A random cloud around zero supports the mean/variance specification.
ggplot(diagnostic_dat, aes(fitted, residual)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_smooth(se = FALSE, color = "steelblue") +
  labs(x = "Fitted value", y = "Residual")

# Departures from the line flag non-normal residual tails or outliers.
ggplot(diagnostic_dat, aes(sample = residual)) +
  stat_qq() +
  stat_qq_line() +
  labs(x = "Theoretical quantile", y = "Sample residual quantile")
```

Diagnostics assess assumptions; they do not prove that a model is correct.

## 9. Labels, facets, and export

```{r}
publication_plot <- ggplot(exam_dat, aes(site, mpg, fill = treatment)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +
  labs(
    title = "Outcome distribution by site and treatment",
    x = "Study site",
    y = "Fuel efficiency (miles per gallon)",
    fill = "Treatment"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom")
publication_plot
```

```{r}
#| eval: false
# Save the plot object explicitly so the exported figure matches the displayed figure.
ggsave("outcome-by-site.png", publication_plot, width = 7, height = 4.5, dpi = 300)
```

Use `facet_wrap(~ variable)` for comparable panels. Keep common axis scales unless different scales are essential and clearly labeled.

## 10. High-yield geometry map

| Question | Geometry |
|---|---|
| Continuous distribution | `geom_histogram()`, `geom_density()` |
| Continuous outcome by group | `geom_boxplot()` with `geom_jitter()` |
| Category count/proportion | `geom_bar()` |
| Precomputed estimate/proportion | `geom_col()` |
| Two continuous variables | `geom_point()` with `geom_smooth()` |
| Repeated trajectory | `geom_line()` with `group = ID` |
| Estimate and interval | `geom_errorbar()`, `geom_pointrange()` |
| Model prediction and band | `geom_line()` with `geom_ribbon()` |
| Multipanel comparison | `facet_wrap()` |
| Residual or Q--Q diagnostic | `geom_point()`, `stat_qq()` |

## 11. Exam references

- 2019 Applied Doctoral Exam, pp. 6 and 8--9: relevant graphical summaries, response proportions, and subtype-overlap panels.
- 2020 Applied Doctoral Exam, pp. 5 and 7: visual evidence and longitudinal descriptive displays.
- 2021 Applied Doctoral Exam, pp. 2 and 9--10: plots supporting assumptions and concise figures for clinical reporting.
- 2023--2025 Applied Doctoral Exams: graphical EDA, predicted trajectories, outcome prevalence by predictor level, and graphical effect evaluation recur throughout.
