---
title: "3. Gaussian covariance models and linear mixed models"
format: html
execute: {echo: true, warning: false, message: false}
---

::: {.callout-note title="Mathematical companions"}
[MVN, ML, REML, and GLS](../math-writeups/01-foundations-mvn-ml-reml.qmd) · [Gaussian mean and covariance models](../math-writeups/03-gaussian-marginal-models.qmd) · [Linear mixed models](../math-writeups/04-linear-mixed-models.qmd) · [Inference and diagnostics](../math-writeups/05-inference-diagnostics.qmd) · [Sample size and power](../math-writeups/13-sample-size-power.qmd)
:::

```{r}
#| label: setup
#| include: false
source("_common.R")
library(dplyr)
library(ggplot2)
library(nlme)
library(glmmTMB)
library(lmerTest)
library(pbkrtest)
library(broom.mixed)
library(influence.ME)
library(lme4)
```

## 1. Aim

Gaussian marginal covariance models estimate population mean trajectories while representing within-subject dependence directly through $\Sigma_i(\theta)$. Linear mixed models represent dependence through latent subject effects and target both population fixed effects and subject-specific predictions. Fixed-effect selection is compared under ML; covariance/random-effect estimation and final inference generally use REML when the fixed-effect design is unchanged.

## 2. Model specification

The marginal model is
$$
Y_i\sim N\{X_i\beta,\Sigma_i(\theta)\}.
$$
Examples include CS, $\Sigma_{jk}=\sigma^2\{(1-\rho)I(j=k)+\rho\}$; discrete AR(1), $\Sigma_{jk}=\sigma^2\rho^{|j-k|}$; continuous AR(1), $\sigma^2\rho^{|t_j-t_k|}$; Toeplitz, $\Sigma_{jk}=\sigma^2\rho_{|j-k|}$; and unstructured $\Sigma$.

The LMM is
$$
Y_i=X_i\beta+Z_i b_i+\varepsilon_i,qquad
b_i\sim N(0,D),\quad \varepsilon_i\sim N(0,R_i),\quad b_i\perp\varepsilon_i,
$$
so $\operatorname{Var}(Y_i)=Z_iDZ_i^{\mathsf T}+R_i$ and
$$
\widehat b_i=DZ_i^{\mathsf T}V_i^{-1}(Y_i-X_i\widehat\beta).
$$

## 3. R packages/functions

`nlme::gls()`/`lme()` are idiomatic when residual covariance structures are primary. `lme4::lmer()` is idiomatic for flexible random-effects models; `lmerTest` and `pbkrtest` add Satterthwaite/Kenward--Roger inference; `broom.mixed` tidies parameters; `influence.ME` supplies cluster-deletion influence diagnostics.

## 4. Data

`nlme::Orthodont` contains four ordered measurements per subject. `visit` indexes equal occasion spacing; `age` is the actual continuous time.

```{r}
data("Orthodont", package = "nlme")
orth <- as.data.frame(Orthodont) |>
  # Always sort within ID before constructing visit order or lagged variables.
  arrange(Subject, age) |>
  group_by(Subject) |>
  # mutate() creates visit order, centered continuous time, and categorical occasion.
  mutate(visit = row_number(), age_c = age - 8, age_f = factor(age)) |>
  ungroup() |>
  mutate(Subject = factor(Subject))
```

## 5. Fit

### Direct residual covariance models

```{r}
# Formula convention: response is left of ~; * includes both main effects and their interaction.
mean_formula <- distance ~ Sex * age_c
# glsControl() changes numerical fitting only: use optim and allow up to 200 iterations.
ctrl <- glsControl(opt = "optim", msMaxIter = 200)

# gls(formula, data, correlation, weights, method): each list item changes only covariance structure.
cov_models <- list(
  independence = gls(mean_formula, orth, method = "REML", control = ctrl),
  CS = gls(mean_formula, orth,
           # form = ~ time | ID tells nlme the within-subject order and clustering.
           correlation = corCompSymm(form = ~visit | Subject),
           method = "REML", control = ctrl),
  AR1 = gls(mean_formula, orth,
            correlation = corAR1(form = ~visit | Subject),
            method = "REML", control = ctrl),
  CAR1 = gls(mean_formula, orth,
             correlation = corCAR1(form = ~age | Subject),
             method = "REML", control = ctrl),
  spatial_exponential = gls(mean_formula, orth,
                            correlation = corExp(form = ~age | Subject, nugget = TRUE),
                            method = "REML", control = ctrl),
  unstructured = gls(mean_formula, orth,
                     correlation = corSymm(form = ~visit | Subject),
                     # varIdent(~1 | age_f) estimates a separate residual variance ratio by age.
                     weights = varIdent(form = ~1 | age_f),
                     method = "REML", control = ctrl)
)

# lapply() extracts fit criteria from every model; do.call(rbind, ...) stacks them by row.
cov_comparison <- do.call(
  rbind,
  lapply(names(cov_models), function(nm) {
    z <- cov_models[[nm]]
    data.frame(model = nm, logLik = as.numeric(logLik(z)), AIC = AIC(z), BIC = BIC(z))
  })
)
cov_comparison[order(cov_comparison$AIC), ]

anova(cov_models$independence, cov_models$CS)  # nested LRT; same fixed-effects design
```

### Toeplitz covariance

`nlme` has no dedicated freely estimated Toeplitz `corStruct`. The exact idiomatic implementation below uses `glmmTMB`; the `nlme::corARMA()` family is stationary but is not, in general, the same parameter space as an unrestricted Toeplitz matrix.

```{r}
# glmmTMB(formula, data, REML): homtoep(0 + occasion | ID) specifies homogeneous Toeplitz covariance.
toeplitz_fit <- glmmTMB::glmmTMB(
  distance ~ Sex * age_c + homtoep(0 + age_f | Subject),
  # dispformula = ~0 avoids adding a second independent residual component.
  dispformula = ~0,
  data = orth,
  REML = TRUE
)
summary(toeplitz_fit)
```

### Fixed-effect and random-effect selection

```{r}
# lmer(formula, data, REML): (age_c | Subject) gives each subject a random intercept and age slope.
# Use REML = FALSE (ML) when comparing fixed-effect formulas.
m_mean_reduced_ml <- lme4::lmer(distance ~ Sex + age_c + (age_c | Subject),
                                data = orth, REML = FALSE)
m_mean_full_ml <- lme4::lmer(distance ~ Sex * age_c + (age_c | Subject),
                             data = orth, REML = FALSE)
anova(m_mean_reduced_ml, m_mean_full_ml)       # fixed effects: ML

# (1 | Subject) is random intercept only; identical fixed effects permit a REML covariance comparison.
m_ri_reml <- lme4::lmer(distance ~ Sex * age_c + (1 | Subject), data = orth, REML = TRUE)
m_rs_reml <- lme4::lmer(distance ~ Sex * age_c + (age_c | Subject), data = orth, REML = TRUE)
AIC(m_ri_reml, m_rs_reml)                      # same fixed effects: REML comparison

final_lmm <- m_rs_reml
# summary(model) reports fixed effects, random-effect covariance, residual SD, and fit diagnostics.
summary(final_lmm)
```

### Multilevel/crossed effects and nonlinear trajectories

```{r}
# Crossed clustering example covered by the multilevel-model extension.
data("Penicillin", package = "lme4")
# Separate (1 | group) terms give crossed random intercepts for plate and sample.
crossed_lmm <- lme4::lmer(diameter ~ 1 + (1 | plate) + (1 | sample), data = Penicillin)
summary(crossed_lmm)

# Nonlinear mixed growth curve: subject-specific asymptote.
data("Loblolly", package = "datasets")
# getInitial() obtains starting values for the named self-starting nonlinear curve.
start_nls <- getInitial(height ~ SSasymp(age, Asym, R0, lrc), data = Loblolly)
# nlme(model, data, fixed, random, start): specify population parameters and subject-varying parameters.
nlmm_growth <- nlme(
  height ~ SSasymp(age, Asym, R0, lrc),
  data = Loblolly,
  # All three curve parameters are fixed; Asym also gets a random intercept within Seed.
  fixed = Asym + R0 + lrc ~ 1,
  random = Asym ~ 1 | Seed,
  start = start_nls
)
summary(nlmm_growth)
```

### Satterthwaite and Kenward--Roger denominator df

```{r}
# lmerTest::lmer() fits the LMM; anova(ddf = ...) chooses the denominator-df approximation.
final_lt <- lmerTest::lmer(
  distance ~ Sex * age_c + (age_c | Subject), data = orth, REML = TRUE
)
anova(final_lt, ddf = "Satterthwaite")
anova(final_lt, ddf = "Kenward-Roger")

kr_full <- lme4::lmer(distance ~ Sex * age_c + (age_c | Subject), orth, REML = TRUE)
kr_reduced <- lme4::lmer(distance ~ Sex + age_c + (age_c | Subject), orth, REML = TRUE)
# KRmodcomp(full, reduced) tests the fixed-effect terms omitted from the reduced model.
pbkrtest::KRmodcomp(kr_full, kr_reduced)
```

### Boundary test for one variance component

```{r}
# The two ML fits differ by a subject random intercept; the null variance lies on the zero boundary.
ri_ml <- lme4::lmer(distance ~ Sex * age_c + (1 | Subject), orth, REML = FALSE)
ols_ml <- lm(distance ~ Sex * age_c, orth)
lr_ri <- max(0, 2 * (as.numeric(logLik(ri_ml)) - as.numeric(logLik(ols_ml))))
p_mixture <- 0.5 * stats::pchisq(lr_ri, df = 1, lower.tail = FALSE)
c(LR = lr_ri, p_50_50_mixture = p_mixture)
```

## 6. Extract & interpret

```{r}
fixef(final_lmm)
as.data.frame(VarCorr(final_lmm))
ranef(final_lmm)$Subject |> head()              # BLUPs/conditional modes
# parm = "theta_" requests variance parameters; method = "profile" uses profile likelihood.
confint(final_lmm, parm = "theta_", method = "profile")

ri_only_icc <- icc_random_intercept(m_ri_reml)
ri_only_icc

# tidy(model, effects, conf.int) returns fixed effects and random-effect SD/correlation in one table.
broom.mixed::tidy(final_lmm, effects = c("fixed", "ran_pars"), conf.int = TRUE)
```

Fixed effects describe the conditional mean at $b_i=0$, which equals the population mean under the identity link. Random-intercept and random-slope variances quantify unexplained between-subject heterogeneity; their correlation describes association between baseline level and change. BLUPs are shrunken subject predictions, not observed subject parameters. The random-intercept ICC is $\tau_0^2/(\tau_0^2+\sigma^2)$; with random slopes, correlation is time-dependent and a single ICC is inadequate. Satterthwaite/KR output reports approximate denominator df; base `lmer()` reports no p-values by design.

## 7. Diagnostics

```{r}
# isSingular() flags an estimated random-effect covariance on/near its boundary.
lme4::isSingular(final_lmm, tol = 1e-4)
final_lmm@optinfo$conv$lme4$messages

# fitted()/resid() create observation-level diagnostic columns from the fitted model.
diag_df <- data.frame(
  fitted = fitted(final_lmm),
  conditional_resid = resid(final_lmm),
  pearson_resid = resid(final_lmm, type = "pearson")
)
# aes(x, y) selects plot axes; geom_hline(yintercept = 0) adds the zero-residual reference.
ggplot(diag_df, aes(fitted, pearson_resid)) + geom_point() + geom_hline(yintercept = 0, lty = 2)

qqnorm(diag_df$conditional_resid); qqline(diag_df$conditional_resid)
re <- ranef(final_lmm)$Subject
qqnorm(re[["(Intercept)"]]); qqline(re[["(Intercept)"]])

# Variogram(model, ~ time | ID) summarizes normalized residual dependence by lag.
vg <- Variogram(cov_models$AR1, form = ~age | Subject, resType = "normalized")
plot(vg, smooth = TRUE)

# influence(model, group) deletes one independent cluster at a time; group names the subject ID.
infl <- influence.ME::influence(final_lmm, group = "Subject")
head(sort(cooks.distance(infl), decreasing = TRUE))
```

For singular or nonconvergent fits: rescale/center time, inspect observations per random coefficient, simplify random-effect correlations with `(1 + age_c || Subject)`, compare optimizers, and retain complexity only when scientifically and empirically supported.

## 8. Caveats

- Compare fixed-effect mean structures under ML. REML likelihoods from different fixed-effect column spaces are not comparable; compare covariance/random-effect structures under REML only with the same fixed effects.
- LRTs are justified for nested models; AIC/BIC may compare nonnested covariance structures but do not account for selection uncertainty.
- For a single variance component under a regular scalar boundary problem, the asymptotic null is $\tfrac12\chi_0^2+\tfrac12\chi_1^2$, not plain $\chi_1^2$. Multiple components or unidentified correlations require a different mixture or parametric bootstrap.
- SAS `PROC MIXED` often requests `DDFM=KR` or `DDFM=SAT`; base `lmer()` supplies neither automatically. `lmerTest` defaults to Satterthwaite unless changed.
- Wald CIs for variance parameters are unreliable near zero; prefer profile-likelihood or bootstrap intervals.
- `nlme::corCAR1()` and `corExp()` use actual time distances; `corAR1()` uses integer occasion gaps.
- `nlme` does not natively expose unrestricted Toeplitz covariance; the `glmmTMB` homogeneous Toeplitz term `homtoep()` is an explicit R-vs-SAS implementation difference. Use `toep()` when occasion-specific marginal variances are required.
- Multilevel random effects require enough independent levels at each hierarchy; a factor with only a few levels should ordinarily be fixed. Nonlinear mixed models are sensitive to starting values and parameterization.
- **Source-backed:** model-building order, covariance choices, REML/ML, boundary tests, BLUPs, and diagnostics follow the source notes. **Standard-practice additions:** `broom.mixed`, `influence.ME`, and `glmmTMB` Toeplitz syntax replace or extend SAS output.

## 9. Source references

- `BIOS-767-Part-007.pdf`, pp. 5--31 and 50--61 (correlated Gaussian model, estimation, covariance structures).
- `BIOS-767-Part-009.pdf`, pp. 4--18, 77--91, and 99--119 (mean/covariance selection; CS, AR, unstructured; ML/REML).
- `BIOS-767-Part-010.pdf`, pp. 5--24, 40--58, and 62--73 (LMM, covariance decomposition, BLUPs, ICC).
- `BIOS-767-Part-011.pdf`, pp. 4--25 and 42--54 (random-effect tests, boundary, fixed-effect inference, KR/Satterthwaite).
- UW STAT571 (LMM, BLUP, ML/REML, inference, serial correlation, diagnostics, and boundary tests).
- BIOS767 Homework 5 (random intercept/slope, BLUP, residual, and influence code).
- `BIOS-767-Part-016.pdf`, pp. 4--23 (multilevel/clustered models); `BIOS_767_Weideman_Typed_Notes_with_Examples.pdf`, p. 90 (nonlinear mixed models).
