---
title: "4. Marginal models and generalized estimating equations"
format: html
execute: {echo: true, warning: false, message: false}
---

::: {.callout-note title="Mathematical companions"}
[Marginal models and GEE](../math-writeups/07-marginal-models-gee.qmd) · [Missing-data weighting](../math-writeups/10-missing-data.qmd) · [Transition models](../math-writeups/11-transition-models.qmd) · [Time-varying covariates](../math-writeups/12-time-varying-covariates.qmd)
:::

```{r}
#| label: setup
#| include: false
source("_common.R")
library(dplyr)
library(ggplot2)
library(geepack)
library(broom)
library(emmeans)
```

## 1. Aim

GEE estimates a population-averaged regression parameter $\beta$ from a correctly specified marginal mean while treating within-cluster association as nuisance. The sandwich covariance gives asymptotically valid inference under working-correlation misspecification when independent clusters are sufficiently numerous and the observation process satisfies the estimating-equation assumptions.

## 2. Model specification

Let
$$
g(\mu_{ij})=x_{ij}^{\mathsf T}\beta,
\qquad \operatorname{Var}(Y_i)=V_i=\phi A_i^{1/2}R_i(\alpha)A_i^{1/2}.
$$
GEE solves
$$
U(\beta)=\sum_{i=1}^m D_i^{\mathsf T}V_i^{-1}(Y_i-\mu_i)=0,
\qquad D_i=\frac{\partial\mu_i}{\partial\beta^{\mathsf T}}.
$$
The model-based covariance is $B^{-1}$; the sandwich covariance is $B^{-1}MB^{-1}$, with $B=\sum D_i^{\mathsf T}V_i^{-1}D_i$ and $M=\sum D_i^{\mathsf T}V_i^{-1}r_ir_i^{\mathsf T}V_i^{-1}D_i$.

## 3. R packages/functions

`geepack::geeglm()` fits GEE with independence (`"independence"`), compound symmetry (`"exchangeable"`), AR(1) (`"ar1"`), and unstructured working correlations (`""unstructured"`); `geepack::QIC()` compares mean/correlation specifications; the internal `geese` object exposes robust and naive covariance and the estimated scale. `gee::gee()` is a useful alternative when SAS-like naive/robust output is desired.

## 4. Data

`geepack::dietox` records pig weight longitudinally by pig (`Pig`), week (`Time`), and feed treatment (`Evit`, `Cu`, `Cu035`).

```{r}
data("dietox", package = "geepack")
dietox <- geepack::dietox |>
  # Pig is the cluster ID; relevel() makes Cu000 the treatment reference category.
  mutate(Pig = factor(Pig), Cu = relevel(factor(Cu), ref = "Cu000")) |>
  # geeglm() expects observations to be ordered by cluster and wave.
  arrange(Pig, Time)
# glimpse(data) compactly checks column names, classes, and representative values.
dplyr::glimpse(dietox)
```

## 5. Fit

```{r}
# Formula convention: Weight is the response; * adds Time, Cu, and the Time-by-Cu interaction.
gee_formula <- Weight ~ Time * Cu

# geeglm(formula, id, waves, data, family, corstr): ID clusters rows; waves orders repeated visits.
gee_ind <- geepack::geeglm(
  gee_formula, id = Pig, waves = Time, data = dietox,
  # family gives response distribution/link; corstr gives the working within-Pig correlation.
  family = gaussian, corstr = "independence", std.err = "san.se"
)
# update() reuses every argument except the named working-correlation choice.
gee_exch <- update(gee_ind, corstr = "exchangeable")
gee_ar1 <- update(gee_ind, corstr = "ar1")
gee_un <- update(gee_ind, corstr = "unstructured")

# QIC(model) supplies GEE model-selection criteria; lapply() evaluates each fitted model.
qic_table <- do.call(rbind, lapply(
  list(independence = gee_ind, exchangeable = gee_exch, AR1 = gee_ar1, unstructured = gee_un),
  function(z) geepack::QIC(z)
))
qic_table

# summary(geeglm_fit) reports coefficients with robust SEs and working-correlation estimates.
summary(gee_ar1)
```

### Time-varying covariate decomposition

Separating a covariate into subject mean and within-subject deviation avoids conflating between- and within-subject associations. Here lagged weight is used only to demonstrate the parameterization; because it is an endogenous response history, causal interpretation requires transition-model assumptions.

```{r}
dietox_tv <- dietox |>
  # group_by(Pig) makes lag() and mean() operate separately within each subject.
  group_by(Pig) |>
  # lag_weight is prior outcome; its Pig mean and deviation separate between/within-subject effects.
  mutate(lag_weight = lag(Weight), mean_lag_weight = mean(lag_weight, na.rm = TRUE),
         dev_lag_weight = lag_weight - mean_lag_weight) |>
  ungroup() |>
  filter(!is.na(lag_weight))

# mean_lag_weight estimates between-Pig association; dev_lag_weight estimates within-Pig association.
gee_within_between <- geepack::geeglm(
  Weight ~ Time + Cu + mean_lag_weight + dev_lag_weight,
  id = Pig, waves = Time, data = dietox_tv,
  family = gaussian, corstr = "exchangeable"
)
# Use summary() to inspect within- and between-subject coefficient estimates and robust SEs.
summary(gee_within_between)
```

QIC may compare working correlations with a common mean model; compare mean models under a common working correlation. A smaller QIC is preferred, but the criterion does not validate the marginal mean.

## 6. Extract & interpret

```{r}
# This helper returns coefficient SEs from robust and model-based GEE covariance matrices.
gee_covariance_table(gee_ar1)

c(
  working_alpha = unname(gee_ar1$geese$alpha),
  scale_phi = unname(gee_ar1$geese$gamma)
)

# tidy(model, conf.int = TRUE) returns one coefficient per row with confidence limits.
broom::tidy(gee_ar1, conf.int = TRUE)

# emtrends(model, specs, var, vcov.): estimate the Time slope by Cu using robust covariance vbeta.
emm_time <- emmeans::emtrends(gee_ar1, ~Cu, var = "Time", vcov. = gee_ar1$geese$vbeta)
emm_time
# pairs() compares Cu-specific slopes; adjust = "holm" controls multiplicity.
pairs(emm_time, adjust = "holm")
```

`Time` is the population-average weekly mean change in the reference feed group; `Time:Cu` terms are population-average differences in slopes. `alpha` parameterizes the chosen working correlation and is not ordinarily a scientific estimand in GEE1. For Gaussian data, $\phi$ estimates residual scale; for binomial/Poisson responses the canonical scale is often fixed at one, although software can estimate a scale depending on options. Robust SEs come from `vbeta`; model-based SEs from `vbeta.naiv` are valid only under the working covariance model.

For nonidentity links, GEE and GLMM coefficients differ. With a logistic random-intercept GLMM,
$$
\beta_{PA}\approx \frac{\beta_{SS}}{\sqrt{1+c^2\sigma_b^2}},
\qquad c=\frac{16\sqrt3}{15\pi}\approx0.588,
$$
so population-averaged log-odds effects are typically attenuated toward zero. Under a probit link the corresponding factor is $(1+z_{ij}^{\mathsf T}Dz_{ij})^{-1/2}$; under the identity link the coefficients coincide.

## 7. Diagnostics

```{r}
# count(Pig) gives cluster sizes; summary() reports their distribution.
cluster_size <- dietox |> count(Pig, name = "n_i")
summary(cluster_size$n_i)

gee_diag <- dietox |>
  # Add fitted means and Pearson residuals to the original observation rows.
  mutate(
    fitted = fitted(gee_ar1),
    pearson = residuals(gee_ar1, type = "pearson")
  )

# Plot residual versus fitted mean; alpha controls point transparency.
ggplot(gee_diag, aes(fitted, pearson)) +
  geom_point(alpha = 0.5) + geom_hline(yintercept = 0, linetype = 2)

# group = Pig connects each subject's residuals; group = 1 asks geom_smooth for one overall trend.
ggplot(gee_diag, aes(Time, pearson, group = Pig)) +
  geom_line(alpha = 0.15) + geom_smooth(aes(group = 1), se = FALSE)

gee_covariance_table(gee_ar1) |>
  mutate(se_ratio = robust_se / model_based_se)
```

Check subject sorting, unique subject-wave keys, cluster-size distribution, leverage from large clusters, mean-model residual trends, and sensitivity of estimates/SEs to plausible working correlations.

## 8. Caveats

- GEE coefficients are population-averaged; GLMM coefficients are subject-specific conditional on random effects. Equality generally fails for nonidentity links.
- Sandwich validity is asymptotic in the number of independent clusters; small-cluster corrections are not the default in `geeglm()`.
- Ordinary GEE is generally valid under MCAR, not arbitrary MAR dropout. Use weighted GEE or another valid missing-data method when observation depends on observed history.
- `geeglm()` requires rows ordered by cluster and, for AR(1), correctly ordered waves.
- Time-varying covariates can be external or endogenous; coefficients for endogenous covariates generally need a conditional/transition interpretation, and naive GEE can be biased when the covariate process depends on prior outcomes.
- QIC is not AIC: it is based on quasi-likelihood and must be interpreted within comparable GEE specifications.
- In `geepack`, `vbeta` is robust and `vbeta.naiv` is model-based. SAS/other R packages may label these `empirical` and `model-based`/`naive`; verify which covariance a printed table uses.
- Scale estimation defaults differ across `geepack`, `gee`, and SAS `PROC GENMOD`/`PROC GEE`; set and report the rule explicitly.
- **Source-backed:** estimating equations, sandwich inference, working correlations, QIC, scale, and PA interpretation follow the notes. **Standard-practice addition:** direct extraction helpers and `emmeans::emtrends()` are idiomatic R replacements for SAS contrasts.

## 9. Source references

- `BIOS-767-Part-014.pdf`, pp. 39--52, 62--68, and 77--80 (GEE, sandwich/model-based covariance, working correlation, QIC).
- UW STAT571 (estimating equations, working covariance, association and scale estimation, robust inference, marginalization, attenuation, and GEE code examples).
- BIOS767 Homework 6 (working-correlation fits, robust/naive SEs, alpha, odds ratios, and Wald tests).
- `BIOS-767-Part-021.pdf`, pp. 3--37 (time-varying covariates and lagged exposure models).
