---
title: "5. GLM review for longitudinal outcomes"
format: html
execute: {echo: true, warning: false, message: false}
---

::: {.callout-note title="Companions"}
[Generalized linear models](../math-writeups/06-generalized-linear-models.qmd) · [Marginal models and GEE](../math-writeups/07-marginal-models-gee.qmd) · [GLMMs and multilevel models](../math-writeups/08-generalized-linear-mixed-multilevel.qmd) · [Ordinal and nominal models](../math-writeups/09-ordinal-nominal-joint-models.qmd) · [Residual diagnostics in R](09-diagnostics-residual.qmd)
:::

```{r}
#| label: setup
#| include: false
source("_common.R")
library(MASS)
library(ordinal)
library(broom)
library(emmeans)
library(dplyr)
library(ggplot2)
```

## 1. Aim

GLMs specify the marginal mean--variance relationship for non-Gaussian responses and provide the mean-model component subsequently embedded in GEE or GLMM. A pooled GLM is a useful benchmark, but its model-based covariance is invalid for repeated observations unless independence is justified or a cluster-robust correction is used.

## 2. Model specification

For an exponential-family response,
$$
f(y;\theta,\phi)=\exp\left\{\frac{y\theta-b(\theta)}{a(\phi)}+c(y,\phi)\right\},
\quad E(Y)=b'(\theta),\quad \operatorname{Var}(Y)=a(\phi)b''(\theta),
$$
and a GLM sets $g(\mu_{ij})=x_{ij}^{\mathsf T}\beta+\log(e_{ij})$, where $e_{ij}$ is exposure for a log-link rate model. Common longitudinal outcomes use binomial-logit, Poisson-log, negative-binomial-log, or Gaussian-identity specifications.

## 3. R packages/functions

`stats::glm()` fits standard GLMs; `MASS::glm.nb()` handles negative-binomial counts; `anova(..., test = "LRT")` compares nested likelihood models; `broom::tidy()` extracts estimates, exponentiated effects, and confidence intervals.

## 4. Data

`MASS::epil` has repeated two-week seizure counts (`y`) for each `subject`, treatment, baseline seizure rate, age, and follow-up `period`.

```{r}
data("epil", package = "MASS")
epil <- MASS::epil |>
  # subject is the repeated-measures ID; period_c sets period 1 as time zero; exposure is time at risk.
  mutate(subject = factor(subject), period_c = period - 1, exposure = 2)
# glimpse(data) compactly checks variable names, classes, and representative values.
dplyr::glimpse(epil)
```

## 5. Fit

```{r}
# glm(formula, family, data): y is response; predictors follow ~; offset(log(exposure)) models a rate.
pois_main <- glm(
  y ~ trt + period_c + lbase + lage + offset(log(exposure)),
  # poisson(link = "log") uses a Poisson count variance and log mean/rate model.
  family = poisson(link = "log"), data = epil
)
# update(model, . ~ . + term) keeps the fitted specification and adds the named interaction.
pois_interaction <- update(pois_main, . ~ . + trt:period_c)

# summary(glm_fit) reports link-scale coefficients, SEs, tests, deviance, and residual df.
summary(pois_interaction)
# anova(..., test = "LRT") compares nested likelihood fits; AIC compares expected predictive fit.
anova(pois_main, pois_interaction, test = "LRT")
AIC(pois_main, pois_interaction)

# Pearson chi-square divided by residual df is a quick overdispersion diagnostic.
dispersion_ratio <- sum(residuals(pois_interaction, type = "pearson")^2) /
  df.residual(pois_interaction)
dispersion_ratio

# Quasi-Poisson inflates SEs; glm.nb() instead estimates a negative-binomial dispersion parameter.
quasi_fit <- update(pois_interaction, family = quasipoisson(link = "log"))
nb_fit <- MASS::glm.nb(
  y ~ trt * period_c + lbase + lage + offset(log(exposure)), data = epil
)
summary(quasi_fit)
summary(nb_fit)
```

### Ordinal response extension

```{r}
epil_ordinal <- epil |>
  # cut() bins y at given breakpoints; ordered() declares increasing response categories.
  mutate(severity = ordered(cut(y, breaks = c(-Inf, 0, 4, Inf),
                                labels = c("none", "moderate", "high"))))

# clmm(formula, data, link): severity is ordinal; (1 | subject) is a subject random intercept.
ordinal_glmm <- ordinal::clmm(
  severity ~ trt + period_c + lbase + (1 | subject),
  # Hess = TRUE retains the Hessian needed for standard errors.
  data = epil_ordinal, link = "logit", Hess = TRUE
)
summary(ordinal_glmm)
```

## 6. Extract & interpret

```{r}
# tidy(model, conf.int, exponentiate): return a coefficient table and convert log effects to ratios.
broom::tidy(pois_interaction, conf.int = TRUE, exponentiate = TRUE)
broom::tidy(nb_fit, conf.int = TRUE, exponentiate = TRUE)

# emtrends(model, specs, var): estimate the period_c trend separately by treatment.
emmeans::emtrends(nb_fit, ~trt, var = "period_c", type = "response")
```

Exponentiated log-link coefficients are conditional rate ratios under the fitted mean model; the offset converts expected counts to rates per exposure unit. An interaction coefficient is a ratio of rate ratios, so constituent main effects are evaluated at `period_c = 0`. The quasi-Poisson fit rescales covariance without defining a full likelihood; the negative-binomial model changes the variance function and supports likelihood-based AIC/LRT.

## 7. Diagnostics

```{r}
# Collect observation-level fitted means, residuals, influence, and subject IDs for diagnostics.
glm_diag <- data.frame(
  fitted = fitted(pois_interaction),
  pearson = residuals(pois_interaction, type = "pearson"),
  deviance = residuals(pois_interaction, type = "deviance"),
  cook = cooks.distance(pois_interaction),
  subject = epil$subject
)

# aes(fitted, pearson) assigns axes; the horizontal zero line reveals systematic residual patterns.
ggplot(glm_diag, aes(fitted, pearson)) + geom_point() + geom_hline(yintercept = 0, lty = 2)
head(glm_diag[order(glm_diag$cook, decreasing = TRUE), ], 8)

c(
  Pearson_dispersion = sum(glm_diag$pearson^2) / df.residual(pois_interaction),
  deviance_dispersion = deviance(pois_interaction) / df.residual(pois_interaction)
)
```

Assess link/mean misspecification, residual dispersion, influential observations/clusters, separation for binary models, and whether an exposure offset is correctly measured.

See [Residual diagnostics for GLMs](09-diagnostics-residual.qmd) for a compact four-panel deviance-residual, Cook's-distance, and leverage workflow.

## 8. Caveats

- These fits ignore repeated-subject correlation; use GEE for population-average inference or GLMM for a conditional random-effects model.
- Do not use a normal Q--Q plot of Pearson or deviance residuals to assess a non-Gaussian GLM; those residuals are not expected to be normal.
- Pearson/deviance dispersion ratios are screens, not formal universal tests, and repeated dependence itself inflates them.
- Quasi-likelihood models have no ordinary likelihood; do not compare their AIC with likelihood fits.
- Exponentiation is link-specific: odds ratios for logit, rate ratios for log, and no ratio interpretation for identity.
- Wald inference may be unreliable with sparse outcomes, boundary estimates, or separation; use likelihood or penalized methods as appropriate.
- Proportional-odds coefficients assume common covariate effects across cumulative logits; assess this constraint before reporting a single ordinal odds ratio. Nominal multinomial models require a separate set of logits and a dependence strategy.
- **Source-backed:** family/link, likelihood, offsets, Wald/LRT, interactions, and overdispersion are in the notes. **Standard-practice addition:** `MASS::glm.nb()` is the idiomatic R extension for overdispersed counts.

## 9. Source references

- `BIOS-767-Part-013.pdf`, pp. 4--21, 26--37, and 60--68 (GLM families, links, inference, fit assessment).
- UW STAT571 (exponential-family models, IRLS/inference, quasi-likelihood, and overdispersion).
- BIOS767 Homework 3 (mean models, robust/model-based extraction, and joint Wald tests).
- `BIOS-767-Part-017.pdf`, pp. 2--26 (ordinal, multinomial, and mixed response types).
