---
title: "8. Multivariate testing and PCA"
format: html
execute: {echo: true, warning: false, message: false}
---

::: {.callout-note title="Mathematical companions"}
[Multivariate-normal foundations](../math-writeups/01-foundations-mvn-ml-reml.qmd) · [Repeated-measures MANOVA](../math-writeups/02-repeated-measures-anova-manova.qmd) · [Ordinal, nominal, and joint responses](../math-writeups/09-ordinal-nominal-joint-models.qmd)
:::

```{r}
#| label: setup
#| include: false
source("_common.R")
library(dplyr)
library(tidyr)
library(ggplot2)
```

## 1. Aim

Classical multivariate tests assess vector mean hypotheses while accounting for covariance among response components. PCA constructs orthogonal linear combinations maximizing sample variance and is used for dimension reduction or visualization; it does not use outcome/group labels and does not itself estimate a longitudinal mean model.

## 2. Model specification

For $p$-variate vectors $Y_i\sim N_p(\mu_g,\Sigma)$, MANOVA tests $H_0:C\mu=0$ using eigenvalues of $E^{-1}H$ (Wilks, Pillai, Hotelling--Lawley, or Roy statistics). PCA diagonalizes the sample covariance/correlation matrix,
$$
S=V\Lambda V^{\mathsf T},\qquad Z_k=(Y-\bar Y) v_k,qquad
\operatorname{Var}(Z_k)=\lambda_k.
$$

## 3. R packages/functions

Base `manova()`/`summary.mlm()` compute multivariate tests; `stats::prcomp()` uses an SVD for stable PCA; `factanal()` fits a common-factor model; `cmdscale()` supplies classical metric MDS/PCoA.

## 4. Data

`nlme::Orthodont` is reshaped to one row per subject and one response column per age, producing a four-dimensional longitudinal response vector.

```{r}
data("Orthodont", package = "nlme")
orth_wide <- as.data.frame(Orthodont) |>
  mutate(age = paste0("age", age)) |>
  select(Subject, Sex, age, distance) |>
  pivot_wider(names_from = age, values_from = distance) |>
  arrange(Subject)
Y <- as.matrix(orth_wide[paste0("age", c(8, 10, 12, 14))])
```

## 5. Fit

```{r}
mv_fit <- manova(Y ~ Sex, data = orth_wide)
summary(mv_fit, test = "Pillai")
summary(mv_fit, test = "Wilks")
summary(mv_fit, test = "Hotelling-Lawley")
summary.aov(mv_fit)                       # componentwise follow-up tests

pca_cov <- prcomp(Y, center = TRUE, scale. = FALSE)
pca_cor <- prcomp(Y, center = TRUE, scale. = TRUE)
summary(pca_cov)
pca_cov$rotation

scree <- data.frame(
  component = seq_along(pca_cov$sdev),
  variance = pca_cov$sdev^2,
  proportion = pca_cov$sdev^2 / sum(pca_cov$sdev^2)
)
ggplot(scree, aes(component, proportion)) + geom_col() +
  scale_x_continuous(breaks = scree$component)

scores <- data.frame(pca_cov$x, Sex = orth_wide$Sex)
ggplot(scores, aes(PC1, PC2, color = Sex)) + geom_point(size = 2)
```

```{r}
# Additional multivariate methods covered in STAT 571.
factor_fit <- factanal(Y, factors = 1, rotation = "none")
factor_fit

pc_dist <- dist(scale(Y))
mds <- cmdscale(pc_dist, k = 2, eig = TRUE)
head(mds$points)
```

## 6. Extract & interpret

Pillai/Wilks tests address the joint sex difference across the four-age response vector; univariate follow-ups require multiplicity control and do not replace the joint test. PCA loadings define each linear combination, scores locate subjects, and $\lambda_k/\sum\lambda_j$ is explained variance. Covariance PCA emphasizes occasions with larger variance; correlation PCA standardizes each occasion. Factor loadings represent a latent common-factor covariance model and are not PCA loadings. PCoA/MDS preserves pairwise distances rather than maximizing variance directly.

## 7. Diagnostics

```{r}
stopifnot(!anyNA(Y), nrow(Y) > ncol(Y))
eigen(cov(Y), symmetric = TRUE)$values

reconstruction_2pc <- sweep(
  pca_cov$x[, 1:2, drop = FALSE] %*% t(pca_cov$rotation[, 1:2, drop = FALSE]),
  2, pca_cov$center, "+"
)
sqrt(mean((Y - reconstruction_2pc)^2))

qqnorm(pca_cov$x[, 1]); qqline(pca_cov$x[, 1])
```

For MANOVA, examine multivariate outliers, covariance heterogeneity, rank, and complete-case support. For PCA, assess scaling, dominance by outliers, scree stability, loading sign arbitrariness, and reconstruction error.

## 8. Caveats

- Multivariate normality and common covariance are finite-sample assumptions for classical MANOVA; Pillai is often comparatively robust but not assumption-free.
- A significant MANOVA does not identify which occasion or contrast drives the result.
- PCA components are sample-dependent, sign-indeterminate, and unsupervised; selecting components by group separation leaks outcome information.
- Complete-case wide matrices can discard substantial longitudinal information; LMM/GEE are preferable for irregular incomplete follow-up when the estimand is a trajectory effect.
- PCA of repeated occasions mixes level and change; interpret loadings against the time ordering.
- **Source-backed:** multivariate test statistics, PCA, factor analysis, PCoA, and MDS follow STAT 571. **Standard-practice addition:** the Orthodont reshaping makes all code self-contained.

## 9. Source references

- UW STAT571 (MVN, Hotelling tests, MANOVA statistics, PCA, factor analysis, PCoA/MDS, and functional PCA).
- `BIOS-767-Part-006.pdf`, pp. 4--23 (longitudinal MANOVA/profile analysis).
